Evaluation of a Cache-Oblivious Data Structure [draft-2]

### 2008-06-01 paper/paper.tex
Spell checked.

\begin{abstract}
In modern computer hardware architecture memory is organized in a hierarchy
consisting of several types of memory with different memory sizes, block
transfer sizes and access times.
Traditionally, data structures are evaluated in a theoretical model that
does not take the existance of a memory hierarchy into account. The
cache-oblivious model has been proposed as a more accurate model.
Although several data structures have been described in this model
relatively little empirical performance data is available.
This paper presents the results of an empirical evaluation of several data
structures in a realistic scenario and aims to provide insight into
the applicability of cache-oblivious data structures in practice.
\end{abstract}

@@

In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious mod\-el is of particular interest. We will then describe what our goals for this research project were and how our work relates to previous research.
A large part of the paper will be dedicated to a description of our research methodology. Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.

\section{Previous Work}
The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.

@@

Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop. In their experimental results the cache-oblivious tree performs worse than the (non-oblivious) alternatives. Nevertheless, they conclude that \textit{``cache-obli\-vious data structures may have significant practical importance''}.

@@

Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-obli\-vious priority deque. Their main result is that although the cache-obli\-vious data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.

@@

Although other behaviour is possible, our queue operates by a first-in,
first-out principle, meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps.

@@

In principle, this also means the operating system has control over when and
which data pages are transfered from main memory to disk storage and back.
However, our experiments (which were performed on a system without a swap file)
were limited to using main memory only.

@@

There are several ways of measuring the time a process takes to execute.
The simplest is counting the number of seconds elapsed since the start of the
program (called \emph{wall clock time}).
This has the disadvantage that concurrently executing processes affect the runtime
of the process being measured, so on a busy system the reported runtime may vary
over different executions.

>>>>

\begin{abstract}
In modern computer hardware architecture memory is organized in a hierarchy
consisting of several types of memory with different memory sizes, block
transfer sizes and access times.
Traditionally, data structures are evaluated in a theoretical model that
does not take the existence of a memory hierarchy into account. The
cache-oblivious model has been proposed as a more accurate model.
Although several data structures have been described in this model
relatively little empirical performance data is available.
This paper presents the results of an empirical evaluation of several data
structures in a realistic scenario and aims to provide insight into
the applicability of cache-oblivious data structures in practice.
\end{abstract}

@@

In the following pages we will give an overview of the different available memory models and explain why the cache-obliv\-i\-ous mod\-el is of particular interest. We will then describe what our goals for this research project were and how our work relates to previous research.
A large part of the paper will be dedicated to a description of our research methodology. Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.

\section{Previous Work}
The cache-oblivious memory model is not the only or the first mod\-el that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.

@@

Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop. In their experimental results the cache-oblivious tree performs worse than the (non-oblivious) alternatives. Nevertheless, they conclude that \textit{``cache-oblivious data structures may have significant practical importance''}.

@@

Olsen and Skov evaluated two cache-oblivious priority queue data structures in practice \cite{olsen2002coa} and designed an optimal cache-oblivious priority deque. Their main result is that although the cache-obliv\-i\-ous data structures they examined make more efficient use of the cache, they do not perform better than traditional priority queue implementations.

@@

Although other behavior is possible, our queue operates by a first-in,
first-out principle, meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps.

@@

In principle, this also means the operating system has control over when and
which data pages are transferred from main memory to disk storage and back.
However, our experiments (which were performed on a system without a swap file)
were limited to using main memory only.

@@

There are several ways of measuring the time a process takes to execute.
The simplest is counting the number of seconds elapsed since the start of the
program (called \emph{wall clock time}).
This has the disadvantage that concurrently executing processes affect the timing
of the process being measured, so on a busy system the reported time may vary
over different executions.
